---
title: "BioStats Project: Dog Breeds"
format: html
editor: visual
execute:
  keep-md: true
---

## Dog Breeds Biostatistical Analysis

Analysis of traits of dogs by breed.

```{r}
#install.packages("tidytuesdayR")
```

```{r}
tuesdata <- tidytuesdayR::tt_load('2022-02-01')
```

```{r}
breed_traits <- tuesdata$breed_traits
```

```{r}
breed_traits <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/breed_traits.csv")
```

```{r}
trait_description <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/trait_description.csv")
```

```{r}
breed_rank_all <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/breed_rank.csv')
```

```{r}
library(tidyverse)
```

```{r}
breed_traits
```

```{r}
trait_description
```

```{r}
breed_rank_all
```

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(breed_traits, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

```{r}
names(breed_rank_all)<-
  janitor::make_clean_names(names(breed_rank_all))
```

```{r}
library (kableExtra)
```

```{r}
library(tidyverse)
```

```{r}
#install.packages("janitor")
names(exploratory_data)<-
  janitor::make_clean_names(names(exploratory_data))
```

```{r}
names(breed_traits)<-
  janitor::make_clean_names(names(breed_traits))
```

# Abstract

The ideas of small dog breeds having "Little Dog Syndrome" and large dog breeds being "Gentle Giants"come mostly from social expectations of these dog breeds. America's favorite large and small dog breeds were used to test these social ideas. The data was consistent with large dogs indeed being "gentle giants." However, the idea of "Little Dog Syndrome" was tested and found not to be shown in the data set as analyzed.

# Introduction

Dogs are often called man's best friend. However, there are just so many dog breeds. The personality of a dog can vary widely based on a number of different factors, including breed. The type of dog preferred by any one person depends on multitudinous factors such as cuteness, size, coat type, and personality traits. A person's opinion of small and large dog breeds commonly comes from a small handful of strongly positive or negative experiences with such dogs. Those who feel strongly about their favorite breeds of dogs may have biases for many reasons, including the social ideas of "Little Dog Syndrome" and larger dogs being considered "Gentle Giants".

"Little Dog Syndrome" is a term used to describe the personality traits of smaller dogs. Often, it is used to accuse small dog breeds of being quite barky, less trainable, not open to strangers, and hyper. The idea of large dogs being "Gentle Giants" is often used by rescue groups who explain that the bad reputation given to large dog breeds comes from bad owners and it is in the nature of large dogs to not act badly. These "Gentle Giants" are more likely to be good with children and other dogs, more open to strangers, and more protective of their homes and families.

# Hypotheses

If there is truth to the idea of “Little Dog Syndrome,” then smaller dog breeds will score unfavorably for trait scores in openness to strangers, trainability, energy level, and barking level.

If there is truth to the social idea of big dogs being “Gentle Giants,” then larger dog breeds will score favorably in good with young children and other dogs, openness to strangers, and watchdog/protective nature.

# Determining the Top 10 Large & Small Dog Breeds

The top 10 Large and Small dog breeds will be selected to narrow the data set. Dogs that are considered medium-sized will be ignored. The rankings from the most recent year included in the data set are what will be considered.

```{r}
breed_rank_all %>%
  select("x2020_rank", breed)
```

```{r}
breed_rank_all %>%
  select("x2020_rank", breed) %>%
  filter(breed == "Retrievers (Labrador)"| breed == "German Shepherd Dogs" | breed == "Retrievers (Golden)" | breed == "Poodles" | breed== "Rottweilers" | breed == "Pointers (German Shorthaired)" | breed == "Australian Shepherds" | breed == "Boxers" | breed == "Great Danes" | breed == "Siberian Huskies")
```

List of top ten large dog breeds to be used for the sake of testing hypothesis: If there is truth to the social idea of big dogs being “Gentle Giants,” then larger dog breeds will score favorably in good with young children and other dogs, openness to strangers, and watchdog/protective nature.

```{r}
breed_rank_all %>%
  select("x2020_rank", breed) %>%
  filter(breed == "French Bulldogs"| breed == "Dachshunds" | breed == "Pembroke Welsh Corgis" | breed == "Yorkshire Terriers" | breed == "Cavalier King Charles Spaniels" | breed == "Miniature Schnauzers" | breed == "Shih Tzu" | breed == "Boston Terriers" | breed == "Pomeranians" | breed == "Havanese")
```

List of top ten small dog breeds to be used to test Hypothesis: If there is truth to the idea of “Little Dog Syndrome,” then smaller dog breeds will score unfavorably for trait scores in openness to strangers, trainability, energy level, and barking level.

# Breed Attributes Analysis Large Dogs

attempt filter command according to slack instructions identify column numbers

```{r}
breed_traits %>%
   select(breed, `watchdog_protective_nature`, `good_with_young_children`, `good_with_other_dogs`, `openness_to_strangers`, `trainability_level`) %>%
  slice(1, 3, 4, 6, 8, 12, 14, 15, 16, 18)
```

Large dogs scored consistently with the idea of large breeds being "Gentle Giants". The selected large dog breeds scored well in every selected trait category with little exception

# Breed Attributes Analysis Small Dogs

```{r}
breed_traits %>%
   select(breed, `energy_level`, `barking_level`, `good_with_other_dogs`, `openness_to_strangers`, `trainability_level`) %>%
  slice(2, 10, 11, 13, 17, 19, 20, 21, 23, 24)
```

The only scores consistent with the idea of "Little Dog Syndrome" are the score trends for energy and barking levels. With a few exceptions the small breeds tend to have a higher energy level and higher barking levels. The small dog breeds scored well with consideration to being good with other dogs and strangers as well as their trainability scores.

# Acknowledgment of Bias

Selecting the top breeds for each size group may prove problematic due to a higher breed rank, most likely being some of the best-behaved of each group. It may be better to select from all across the breed rank list or perhaps also look at the lowest scorers in terms of rank to see if they also following the trends. Selecting the breeds to be used based off of how they ranked could bias the data in favor of the dog sizes being better behaved than is truthful for the trend of such breeds.
