---
title: "BioStats Project: Dog Breeds"
format: html
editor: visual
execute:
  keep-md: true
---

## Dog Breeds Biostatistical Analysis

Analysis of traits of dogs by breed.

```{r}
#install.packages("tidytuesdayR")
```

```{r}
tuesdata <- tidytuesdayR::tt_load('2022-02-01')
```

```{r}
breed_traits <- tuesdata$breed_traits
```

```{r}
breed_traits <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/breed_traits.csv")
```

```{r}
trait_description <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/trait_description.csv")
```

```{r}
breed_rank_all <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-02-01/breed_rank.csv')
```

```{r}
breed_traits
```

```{r}
trait_description
```

```{r}
breed_rank_all
```

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(breed_traits, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

# Hypotheses

If there is truth to the idea of “Little Dog Syndrome,” then smaller dog breeds will score unfavorably for trait scores in openness to strangers, trainability, energy level, and barking level.

If there is truth to the social idea of big dogs being “Gentle Giants,” then larger dog breeds will score favorably in good with young children and other dogs, openness to strangers, and watchdog/protective nature.
